System for activity tracking of patients with kidney failure

### 2015-06-05 MScCS.tex
Versionen av rapporten innan redovisningen. Mao den version som var innan vi började göra de ändringar som Görel och opponenter pekade på.

%todo = interact to interaction in fig 6.4
\documentclass{cslthse-msc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{tikz}

@@

%\title{RSAT}
%\subtitle{A system for activity tracking of patients with kidney failure.}
%\company{The Corporation AB LTD Inc}
\supervisors{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}{Don Jeer, \href{mailto:djeer@xy.lth.se}{\texttt{djeer@xy.lth.se}}}
\examiner{Jane Doe, \href{mailto:jane.doe@cs.lth.se}{\texttt{jane.doe@cs.lth.se}}}

@@

\acknowledgements{
If you want to thank people, do it here, on a separate right-hand page. Both the U.S. \textit{acknowledgments} and the British \textit{acknowledgements} spellings are acceptable.

We would like to thank Lennart Andersson for his feedback on this template.
}

\theabstract{
Many people suffering from kidney failure are in need of a kidney transplantation. A problem in the health care is that these patients cannot undergo surgery if they have too much belly fat. Regular exercise and physical activity are therefore crucial for this group of people. A project group at the Skåne University Hospital has recently been established to help patients with kidney failure to lose weight. A question they asked themselves was whether it was possible to use activity tracking devices to monitor patient’s activities.

The purpose of this thesis was to design and evaluate a system that can be used to track the physical activities of the patients. The system was built using a Sony SmartBand, a wristband collecting and analyzing data about a user's daily activities. The conclusions of the thesis is that the system can be built but becomes complicated to use due to Sony’s API.
}

\keywords{kidney failure, activity tracking, loose weight, Sony Lifelog}

@@

Many people suffering from kidney failure are in need of a kidney transplant. A problem in the health care is that patients can not undergo surgery if they have too much belly fat. It is very hard for this group of people to loose weight, as further described in section \ref{sec:kiney}. Regular exercise and physical activity are therefore crucial for these patients.

This thesis presents a system with the aim of helping over-weight patients with kidney failure to loose weight. The system is called the Region Skåne Activity tracking, RSAT. 
itACiH is a research project that investigates the possibilities of advanced health care at home. As a part of this project a planning-system was developed that is currently being used at the hospitals in Skåne. The planning system is built as a web application and can fetch data from sensors that are installed in a patient's home. RSAT is thought to be integrated with the planning system. 
%% Förslag på ändring
% RSAT was developed to be integrated in a planning system already in use at the hospital. The planning system is develop in a research project called itACiH. itACiH investigates the possibilities of advanced health care at home. The planning system is built as a web application and can fetch data from sensors that are installed in a patient's home. 
A project group at the Skåne University Hospital (called the RSAT-group in this thesis) has recently been established to help patients in need of a kidney transplant to loose weight. One of their wishes was to use activity tracking devices in the project.

Sony has developed wearable devices that can track a user’s daily activities. The data collected by these devices are saved online in their database and accessible through an API called the Lifelog API. Sony wants to investigate what different use cases that exists for their wearable devices and has chosen to sponsor this project with smart wristbands for the patients to wear.

@@

%The purpose of this thesis is to build a system that can be used in the health care to help patients with kidney failure to loose weight.
The purpose of this thesis was to design and evaluate a system that can be used to track the physical activities of the patients. The Sony Lifelog API will be used as the foundation for the system. As a step in building the system, an investigation about how well suited the API is to be used will also be performed.

The following questions will be %used to fulfill the purpose: 
answered in this thesis:

@@

\begin{quote}
\centering
\emph{How well suited is RSAT, the system to be developed, to be used in the health care?}
\end{quote}
%Frågeställningarna behöver vi tyvärr fortfarande fila på. vettitusan hur vi ska göra här...

@@

\end{comment}
\section{Method}
Before the development of RSAT could start, an interview with the RSAT-group was held. Their wishes was used as input when the requirements for the system was created. Thereafter a proof-of-concept system was developed to verify that it was possible to use the Lifelog API.

Then the phase of writing production code for RSAT begun. The approach for the development was to use iterations. Each iteration resulted in a demo for the RSAT-group. During the demo feedback was given from the RSAT-group. The experience from the development of the system created the foundation for the evaluation of both the Lifelog platform and RSAT.

Each iteration followed these steps:

\begin{enumerate}
\item Identify the requirements for the current prototype.
\item Implement functionality in the system such that the system fulfills the requirements.
\item Test the developed prototype and publish it such that others can test our software.
\item Receive feedback from testers and use it as input to the next prototype. 
\item Receive new requirements on the system and if the requirements are feasible, add them to a later iteration. 
\end{enumerate}

@@

\section{Disposition}

The first part of this thesis gives an overview of the landscape that RSAT acts in. First a medical background is given, then a technical background describing the techniques used in RSAT. After that related work describing  different warble technologies, platforms for aggregation of activity data and  systems similar to RSAT.

The second part of this thesis contains a description of the development of RSAT. From the initial thoughts in the RSAT group to the last prototype. Telling about all decisions and struggles.

At last the RSAT and Lifelog is evaluated and compared to other systems described in the first part of the thesis. Also some conclusions are drawn.

@@

\section{Chronic conditions}

@@

As seen in figure \ref{fig:chronic-fig} physical inactivity is one factor that could lead to a chronic condition.

%There are also other chronic condition that will benefit from physical activity and more information about one of them in the section below.

%% Den utmarkerade texten ovan kanske ska mergeas, läggas till igen på något vis när kidney failure-sektionen är genomgången.

@@

People with severe CKD are in need of a kidney transplant. Until a transplant can be performed, dialysis is required. Without dialysis, the patients would have a poor life expectancy\cite{KIDNEYFOUNDATION}. Dialysis is a cumbersome treatment process for a patient as it takes up a lot of time and adds a lot of constraints to the patient’s life. Dialysis is also expensive for the society. A rough estimate of the annual cost of treating one patient with dialysis in Sweden is 500 000 – 1 000 000 SEK.

@@

The response to a valid request made to the API contains a result-section. This result-section is a set of data represented as JSON objects. If the requested data is too large to fit in the size of one response message, the data is split up into several messages and a link-field is inserted into the response. This link-field contains a link pointing to the location from where the rest of the data can be fetched.

@@

\begin{enumerate}
	\item A client requests authentication from a resource owner. 
	\item The client receives an authorization grant credential from the resource owner.    
	\item The client requests an access token from the authorization server. The server verifies the authorization grant credential and returns an access token. 
	\item The client receives an access token, which he/she can use to access the protected data in the resource server. 
\end{enumerate}

The Authorization Grant is a credential representing the resource owner's authorization. The IETF's OAuth 2.0 specification defines four different grant types; authorization code, implicit, resource owner password credentials and client credentials. The Sony Lifelog API uses the authorization code as its grant type and the other types will therefore not be further discussed\cite{LifeLogAuth}. When the authorization code grant type is used, the resource owner is redirected to an authentication server when the client requests authentication. When the resource owner has granted the request, the resource owner is redirected back to the client together with the authorization code, which is appended in the URL. The authentication code is short lived and single-use, meaning that if the code is used more than once, the access tokens related to the code will be revoked.

@@

\begin{itemize}
\item \emph{PalCom service} - A PalCom service is a type of self-contained system. The service contains logic for both computations and interactions with the outside world. The implementation of services is made in Java and can therefore leverage the whole Java ecosystem. Every version of a service has unique id called \emph{serviceId} and every instance of a service has an unique id called \emph{instance}.

@@

\item \emph{PalCom assemblies} - PalCom Assemblies are rules explaining how PalCom commands shall be routed in a PalCom network. An assembly connects a service out-commands with another service in-commands and vice-versa. This is what makes it possible for different services to communicate with each other. An assembly connecting services with each other could have the following rule:  “when service s1 sends command cmdOut, send it to cmdIn on service s2". Assemblies are meant to make the interconnection between different services easy to implement.

@@

\item \emph{PalCom Runtime Component} - Each service has a mechanism, which makes it possible to discover other services and to advertise its existence. This mechanism is called the PalCom Runtime Component. Different services can run on an application called theThing. theThing is a type of PalCom device. It has a JVM for running services and an interpreter for assemblies. It also handles all the network configuration and connections.

@@

There exists a wide range of different fitness tracking products on the market. Polar, Fitbit, Garmin, Samsung, to name a few, are all companies, which have released one or several fitness devices. Many of the products have more or less the same functionality. Basic features they have in common are presented in section \ref{sec:Wearable}. The data gathered by the fitness devices needs to be stored somewhere. Three platforms that aggregates different health and fitness data are therefore also presented in this section.

@@

Movement tracking and sleep tracking are two common features in the smart wristbands that exist on the market. According to Quantified Wellness, an article published by RGA Reinsurance Company\cite{callaway2015quantified}, most of the data are collected through an accelerometer in the device. Many of the smart wristbands also takes advantages of the GPS in the smartphone they are connected to in order to increase the accuracy of the measurements.

Other features that exist in wristbands on the market are the ability to track a user's heart rate, body temperature, blood oxygen level, respiration rate and galvanic skin response. The ways these sensors are used in different wristbands differ. In some wristbands, the user must activate a sensor in order to make a measurement and in some other wristbands data are collected without any interaction with the user at all.

The most common type of heart rate sensor is optical and uses LED-lights and an optical sensor to detect the amounts of blood flowing through the wrist in order to calculate a user’s pulse. The accuracy of different optical heart rate sensors varies a lot. In a test performed by CNET, four fitness trackers with an optical heart rate sensor were used to measure the heart rate. The results were then compared to the pulse measured by an EKG machine. The variance in the results were big.  At a pulse rate of 80 – 90 bpm, all of the devices was able to measure the heartbeat and was at the most 10\% off from the real heart rate. At a heart rate of 160 – 170 bpm, one of the trackers could not measure any data at all and two of the trackers were more than 50\% off\cite{CNET}. One of the devices used in the test, a Samsung Galaxy S5 smartphone, proved to be very accurate. It was at most 3\% off compared to the EKG-result.

A chest strap was also used to measure the pulse. The results from the chest strap was very close to the EKG when the heart rate was between 160-170 and about 10\% off when it measured the heart rate at 80 – 90 bpm. The chest strap measures the pulse by measuring electrical pulse in a similar way to the EKG-machine and it is therefore expected that the results from the chest strap should be more accurate.

Another type of heart rate monitor used in some wristbands is an impedance sensor. The impedance sensor measures the resistance of body tissue and uses that data to calculate the pulse. Jawbone Up3 and Jawbone Up4 are using this technique. These devices have recently hit the market and any comments about their actual accuracy can therefore not be given. The pro with the impedance sensor is that it can be used to measure more data such as respiration rate and galvanic skin response. Important to mention is that Up3 and Up4 will mainly focus on resting heart rate. This is something that most of the manufacturers of fitness trackers that are using optical sensors also state that they do.

@@

The awareness and usage of wearables are on the rise. However, it has been showed that a third of all users who obtained a fitness device more than 12 months ago are no longer using it. The data collected by different fitness devices has also been showen to differ a lot. For example a equal long walk could differ between different activity tracker from 8.9 km to 14.5 km \cite{wearable-technology-cannot-be-trusted}.

@@

\subsubsection{Google Fit}
Google Fit is an open platform for both Android and web applications to share and store fitness data. An application collecting fitness data from a fitness tracker can store the data in the Google Fit API. It can also read other data from the API and use it for different calculations. This makes it simple for an application to integrate fitness data collected by different fitness tracking devices. It also makes it convenient for a user to change fitness tracking device as the data the device is collecting will have the same structure in the API\cite{GoogleFitOverview}.

@@

In the Health application the user explicitly controls which apps that are allowed to access the HealthKit store and what data they are allowed to read or write. When an app request data from Apple Health and is not granted access it gets back an empty response. This means that an application, which is denied access to some data cannot know whether it has been denied or if the data does not exist.

@@

Developers can read and write data to the endpoints in the API. The API defines what data types that can be stored in the platform. The communication to the API is REST-based. The endpoints contain different data such as information about a user's body temperature and heart rate. Third party applications can also create custom data entries using the \verb!generic_event! endpoint.\cite{JawboneCustomEndpoint}

@@

The three companies mentioned above have all built their wellness programs around a common portal with a dashboard where all the participating employees can view their progress and compare it with their colleagues. The employees can also be organized into teams and specific team goals can be set for the teams. Administrators of a wellness program can measure the progress of the employees and their amounts of healthy activities through an administrative dashboard where statistics and summaries about the programs can be generated and viewed.

@@

The use of smart band in wellness programs is a fairly modern phenomenon as seen in figure \ref{fig:timline}\cite{fitbitWellnessStat}\cite{jawboneWellnessStat}\cite{garminWellnessStat}.

\begin{figure}[!hbt]
\centering
\begin{chronology}[1]{2012}{2016}{10cm}[10cm]
\event{\decimaldate{20}{12}{2013}}{Firbit}
\event{\decimaldate{12}{10}{2014}}{Jawbone}
\event{\decimaldate{5}{1}{2015}}{Garmin}
\end{chronology}
\caption{A timeline showing when three different smart wear manufacturers released their wellness programs.}
\label{fig:timline}
\end{figure}

@@

Fruit Street is a software system built to help health care providers connect with their patients. The system is made for the American market and has different features. Fruit Street makes it possible for the clinicians to have video chats and screen sharing sessions with their patients. Appointments can be booked and payment collected. Patients can upload pictures of what food they are eating and share it with the clinicians for nutrition purposes.

@@

The user-interface to be used by the physiotherapist should be easy to use. It was also expressed that there was a need for at least two kinds of views: (1) - An overview, where all patients can be listed and where it is easy to see the status of each patient. (2) - A specific view where details about each patient can be viewed.

@@

Two things had to be taken into account when the requirements for RSAT were made. (1) – Sony Lifelog was already the chosen data source. This meant that only data that could be tracked by Lifelog devices and that was accessible via the Lifelog API could be fetched into RSAT.  (2) – RSAT was to be integrated with a system that already existed (the itACiH planning system).  This meant that RSAT had to be architecture such that integration was possible..

@@

\section{Requirements}
\label{sec:req}

Här beskriv krav. Vi har delat upp de i kategorier...

@@

It was also decided that it should be possible to browse the history of all the collected data in RSAT. This function was considered important, since it would make it possible to see if the patient’s make any progression or follow the instructions.

@@

\begin{itemize}
    \item RSAT shall be built as a system the can be reached from the Internet.
    \item The user-interface shall be easy to use.
    \item The data in RSAT shall be easy to access for the staff using it.
    \item It shall be possible to create specific goals for each patient in the system.
    \item It shall be possible to create goal for each datatype that exists in RSAT.
    \item It shall be possible for the clinician to have an overview of all patients, showing if they have completed their goals or not.

@@

\subsection{Other requirements}
\begin{itemize}
    \item Each patient shall be able to view the goals that the clinician has created for him/her.
    \item It shall be possible to see if a patient has been inactive for too long.% Inactive means that the patient has not performed any physical activity during a time interval that is considered to be too long.
    \item It shall be easy in later stages to add wristband from another manufactures. 
    \item A device/set of devices (smartphone and wristband) used by a patient shall be reusable. This means that a new patient can be equipped with the device/set of devices after the first patient has finished the program. 
\end{itemize}

\section{Initial idea}
The initial idea was to develop a specific Android application that was installed on each patient’s smartphone. This would have been the ultimate choice but this idea was thrown and it was chosen that the patients would use the Sony Lifelog Application instead. The two major reasons for this decision were. 
(1) Sony has invested a lot of resources to create the Lifelog application that runs on most newer Android devices. This application has almost all the functionality that the physiotherapist initially asked for.
(2) That the schedule for building the RSAT was quite tight. If an Android application should be developed the time frame would have to be changed. It would also be hard to create an application that the patient’s would prefer to use instead of the Lifelog application.

The Sony Lifelog API is a read-only API. Hence, it is not possible to write any data to the API. This means that it is not possible for the physiotherapist to set goals that are automatically updated in the patients Lifelog accounts. What also is a problem is that the goals are not accessible at all through the API. This means that in order to set goals, the physiotherapist must have a meeting with the patient where they together set the different goals for the patient. At this meeting, the physiotherapist inserts the goals in RSAT and the patient inserts the goals in the Lifelog application.

\section{Final idea}
\label{sec:proposed-solution}
% mer lim
It was decided that the patient will use the lifelog application and that the system developed in the thesis only is a web application. 
Figure \ref{fig:pro} shows an overview of the solution followed by a description of the steps for transferring the data from the patient to the physiotherapist.

@@

\begin{enumerate}
    \item Each patient added in RSAT has a smart Sony SmartBand wristband around his or her wrist.
    \item The wristband collects data about the patient movements and the data is categorized into different activities.
    \item A Sony smartphone collects the data from the wristband via Bluetooth 4.0. 
    \item The data is stored in the Sony Lifelog application on the smartphone and synchronized to Sony's cloud.
    \item RSAT fetches the data for a patients account, analyses it, and saves it to a database.
\item The RSAT web application retrieves the data from all patients in the database and visualizes it to the clinicians.
\item From the computer the nurse can interact with RSAT and see the progress of the patients. 
\end{enumerate}

@@

Each patient in the program needs an activity tracker and something to sync the data to the cloud with. Because of the use of Sony smart band SWR10 and limitations of Lifelog to only track biking on SONY phones, it was decided to equip every patient with a SONY phone and SWR10. The patient borrowed these phones and smart band from the hospital. To make it easy for the clinicians and the patients to get started , each phone and smart band was preconfigured using the following steps:

\begin{enumerate}
%    \item{Phone and SmartBand is unboxed and charged.}
    \item{A new Google account is created for the phone.}
    \item{The Google account is used to install the needed apps:
%\begin{itemize}
%    \item Lifelog
 %   \item SmartBand SWR10
%    \item SmartConnect 
%\end{itemize}
}
    \item{The Lifelog application is logged in via the Google account.}
   \item{A SWR10 is connected with the phone.}
    \item{The phone is added to our system and given a name.}
    \item{The phone and the SmartBand is marked with the name of the phone and the SmartBand inserted in the system.}
\end{enumerate}

When the clinician gives the smart band and the phone to the patient they only need to link the patients profile to the phone in RSAT. %%Denna meningen är urusel

\chapter{Development}
This chapter describes the development cycles of the RSAT. It also gives a brief introduction to the different 3rd-party libraries and PalCom services that have been used to build the RSAT.

\section{3rd-party libraries and services}
%This section gives a brief introduction to the different 3rd-party libraries and PalCom services that have been used to build the RSAT.

@@

@@

>>>>

%todo = interact to interaction in fig 6.4
\documentclass{cslthse-msc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{pdfpages}

@@

%\title{RSAT}
%\subtitle{A system for activity tracking of patients with kidney failure.}
%\company{The Corporation AB LTD Inc}
\supervisors{Boris Magnusson, \href{mailto:Boris.Magnusson@cs.lth.se}{\texttt{Boris.Magnusson@cs.lth.se}}}{Fredrik Nilsask, \href{mailto:fredrik.nilsask@cs.lth.se}{\texttt{fredrik.nilsask@cs.lth.se}}}
\examiner{Görel Hedin, \href{mailto:gorel.hedin@cs.lth.se}{\texttt{gorel.hedin@cs.lth.se}}}

@@

\acknowledgements{
TODO!
%If you want to thank people, do it here, on a separate right-hand page. Both the U.S. \textit{acknowledgments} and the British \textit{acknowledgements} spellings are acceptable.

%We would like to thank Lennart Andersson for his feedback on this template.
}

\theabstract{
Many people suffering from kidney failure are in need of a kidney transplant. A problem in the health care is that patients cannot undergo surgery if they have too much belly fat. Regular exercise and physical activity are therefore crucial for this group of people. A project group at the Skåne University Hospital has recently been established to help patients with kidney failure to lose weight. A question they asked themselves was whether it was possible to use activity tracking devices in the project. The purpose of this thesis is to design and evaluate a system that can be used to track the physical activities of the patients. The system will be built using a Sony SmartBand, a wristband collecting and analyzing data about a user's daily activities.
}

\keywords{kidney failure, activity tracking, lose weight, Sony Lifelog}

@@

Many people suffering from kidney failure are in need of a kidney transplant. A problem in the health care is that patients cannot undergo surgery if they have too much belly fat. It is very hard for this group of people to lose weight, as further described in section \ref{sec:kiney}. Regular exercise and physical activity are therefore crucial for these patients.

This thesis presents a system with the aim of helping over-weight patients with kidney failure to lose weight. The system is called the Region Skåne Activity Tracking, RSAT. 
itACiH is a research project that investigates the possibilities of advanced health care at home. As a part of the itACiH project a planning-system was developed that is currently being used at the hospitals in Skåne. The planning system is built as a web application and can fetch data from sensors that are installed in a patient's home. RSAT is thought to be integrated with the planning system. 
%% Förslag på ändring
% RSAT was developed to be integrated in a planning system already in use at the hospital. The planning system is develop in a research project called itACiH. itACiH investigates the possibilities of advanced health care at home. The planning system is built as a web application and can fetch data from sensors that are installed in a patient's home. 
A project group at the Skåne University Hospital (called the RSAT-group in this thesis) has recently been established to help patients in need of a kidney transplant to lose weight. One of their wishes was to use activity tracking devices in the project.

Sony has developed wearable devices that can track a user’s daily activities. The data collected by these devices are saved online in their cloud and accessible through an API called the Lifelog API. Sony wants to investigate what different use cases that exists for their wearable devices and has chosen to sponsor this project with smart wristbands for the patients to wear.

@@

The purpose of this thesis is to build a system that can be used in the health care to help patients with kidney failure to lose weight. The Sony Lifelog API will be used as the foundation for the system. As a step in building the system, an investigation about how well suited the API is to be used will be performed.

The following questions will be used to fulfill the purpose:

@@

\begin{quote}
\centering
\emph{How well suited is RSAT, the system to be developed, to be used in the health care?}
\end{quote}

@@

\end{comment}
\section{Method}
Before the development of RSAT could start, an interview with the RSAT-group was held. Their wishes was used as input when the requirements was created. Thereafter a proof-of-concept system was developed to verify that it was possible to use the Lifelog API.

Iterative prototyping was used to develop RSAT. For each iteration, the current requirements was identified and implemented. Each iteration resulted in a demo that the RSAT-group evaluated. Feedback from the RSAT-group was used as input to the next iteration. Sometimes the feedback resulted in new requirements on the system and if those was feasible, they was added to a development iteration.

In parallel with the development, a research investigating different wearable devices and different platforms used for activity tracking was performed. The experience from this investigation in combination with the experience from the development of the system created the foundation for the evaluation of both the Lifelog platform and RSAT.

@@

\section{Disposition}
The thesis is organized into three different parts; an overview part, a development part and an evaluation part.

The overview part consists of chapters 2, 3 and 4. It gives a medical background, describing why RSAT is needed and a technical background that describes the Lifelog platform. It also contains an introduction to PalCom, the framework that RSAT is built on. The last chapter in this part describes related work. Related work includes information about different wearable technologies, platforms for aggregation of activity data and other systems that are similar to RSAT.

The development part consists of chapters 5 and 6. This part contains information about the requirements on the system, what decisions that were made, why the decisions were made and how RSAT was implemented.

The evaluation part consists of chapters 7 and 8. RSAT and Lifelog are evaluated and compared to the other systems described in the first part of the thesis.

@@

\section{Chronic conditions}
\label{sec:chronic}

@@

Physical inactivity, as seen in figure \ref{fig:chronic-fig} , is one factor that could lead to a chronic condition.

@@

People with severe CKD are in need of a kidney transplant. Until a transplant can be performed, dialysis is required. Without dialysis, the patients would have a poor life expectancy\cite{KIDNEYFOUNDATION}. Dialysis is a cumbersome treatment process for a patient as it takes up a lot of time and adds a lot of constraints to the patient’s life. Dialysis is also expensive for the society. A rough estimate of the annual cost of treating one patient with dialysis in Sweden is 250 000 – 500 000 SEK. \cite{sr}

@@

The response to a valid request made to the API contains a set of JSON objects. If the requested data is too large to fit in the size of one response message, the data is split up into several messages and a link-field is inserted into the response. This link-field contains a link pointing to the location from where the rest of the data can be fetched.

@@

\begin{enumerate}
	\item A client requests authentication from a resource owner. 
	\item The client receives an authorization grant credential from the resource owner.    
	\item The client requests an access token from the authorization server. The server verifies the authorization grant credential and returns an access token. 
	\item The client receives an access token, which can use to access the protected data in the resource server. 
\end{enumerate}

The Authorization Grant is a credential representing the resource owner's authorization. The IETF's OAuth 2.0 specification defines four different grant types; authorization code, implicit, resource owner password credentials and client credentials. The Sony Lifelog API uses the authorization code as its grant type and the other types will therefore not be further discussed\cite{LifeLogAuth}. When the authorization code grant type is used, the resource owner is redirected to an authentication server when the client requests authentication. When the resource owner has granted the request, the resource owner is redirected back to the client together with the authorization code, which is appended in the URL. The authentication code is short lived and single-use.

@@

\begin{itemize}
\item \emph{PalCom service} - A PalCom service is a type of self-contained system. The service contains logic for both computations and interactions with the outside world. The implementation of services is made in Java and can therefore leverage the Java ecosystem. Every version of a service has a unique id called \emph{serviceId} and every instance of a service has an unique id called \emph{instance}.

@@

\item \emph{PalCom assemblies} - PalCom Assemblies are rules explaining how PalCom commands shall be routed in a PalCom network. An assembly connects a service out-commands with another service in-commands and vice-versa. This is what makes it possible for different services to communicate with each other. An assembly connecting services with each other could have the following rule:  “when service s1 sends command cmdOut, send it to command cmdIn on service s2". Assemblies are meant to make the interconnection between different services easy to implement.

@@

\item \emph{PalCom Runtime Component} - Each service has a mechanism, which makes it possible to discover other services and to advertise its existence. This mechanism is called the PalCom Runtime Component. Different services can run on an application called theThing. theThing is a type of PalCom device. It has a JVM for running services and an interpreter for assemblies. It also handles all the network configuration and connections in a PalCom network.

@@

There exists a wide range of different fitness tracking products on the market. Polar, Fitbit, Garmin, Samsung, to name a few, are all companies, which have released one or several fitness devices. Many of the products have more or less the same functionality. Basic features they have in common are presented in section \ref{sec:Wearable}. The data gathered by fitness devices needs to be stored somewhere. Three platforms that aggregates different health and fitness data are therefore also presented in this section.

@@

Movement tracking and sleep tracking are two common features in the smart wristbands that exist on the market. According to Quantified Wellness, an article published by RGA Reinsurance Company\cite{callaway2015quantified}, most of the data are collected through an accelerometer in the device. Many of the smart wristbands also takes advantages of the GPS in the smartphone they are connected to and thereby increases the accuracy of the measurements. Data collected in this type of fitness devices has shown to differ a lot. For example an equally long walk could differ between different activity tracker from 8.9 km to 14.5 km \cite{wearable-technology-cannot-be-trusted}.

Other features that exist in wristbands on the market are the ability to track a user's heart rate, body temperature, blood oxygen level, respiration rate and galvanic skin response. The ways these sensors are used in different wristbands differ. In some wristbands, the user must activate a sensor in order to make a measurement and in some wristbands data are collected without any interaction with the user at all.

The most common type of heart rate sensor is optical and uses LED-lights in combination with an optical sensor to detect the amounts of blood flowing through the wrist in order to calculate a user’s pulse. The accuracy of different optical heart rate sensors varies a lot. In a test performed by CNET, four fitness trackers with an optical heart rate sensor were used to measure the heart rate. The results were then compared to the pulse measured by an EKG machine. The variance in the results were big.  At a pulse rate of 80 – 90 bpm, all of the devices was able to measure the heartbeat and was at the most 10\% off from the real heart rate. At a heart rate of 160 – 170 bpm, one of the trackers could not measure any data at all and two of the trackers were more than 50\% off\cite{CNET}. One of the devices used in the test, a Samsung Galaxy S5 smartphone, proved to be very accurate. It was at most 3\% off compared to the EKG-result. A chest strap was also used to measure the pulse. The results from the chest strap was very close to the EKG when the heart rate was between 160-170 and about 10\% off when it measured the heart rate at 80 – 90 bpm. The chest strap measures the pulse by measuring electrical pulse in a similar way to the EKG-machine and it is therefore expected that the results from the chest strap should be more accurate.

Another type of heart rate monitor used in some wristbands is an impedance sensor. The impedance sensor measures the resistance of body tissue and uses that data to calculate the pulse. Jawbone Up3 is using this technique\cite{up3}. These devices have recently hit the market and any comments about their actual accuracy can therefore not be given. The pro with the impedance sensor is that it can be used to measure more data such as respiration rate and galvanic skin response. Important to mention is that Up3 and Up4 will mainly focus on resting heart rate. This is something that most of the manufacturers of fitness trackers that are using optical sensors also state that they do.

@@

The awareness and usage of wearables are on the rise. However, it has been showed that a third of all users who obtained a fitness device more than 12 months ago are no longer using it.

@@

\subsubsection{Google Fit}
Google Fit is an open platform for both Android and web applications to share and store fitness data. An application collecting fitness data from a fitness tracker can store the data in Google Fit via the API. It can also read other data from the API and use it for different calculations. This makes it simple for an application to integrate fitness data collected by different fitness tracking devices. It also makes it convenient for a user to change fitness tracking device as the data the device is collecting will have the same structure in the API\cite{GoogleFitOverview}.

@@

In the Health application the user explicitly controls which apps that are allowed to access the HealthKit store and what data they are allowed to read or write. When an application requests data from Apple Health and is not granted access it gets back an empty response. This means that an application, which is denied access to some data cannot know whether it has been denied or if the data does not exist.

@@

Developers can read and write data to the endpoints in the API. The API defines what data types that can be stored in the platform. The communication to the API is REST-based. The endpoints contains different data such as information about a user's body temperature and heart rate. Third party applications can also create custom data entries using the \verb!generic_event! endpoint.\cite{JawboneCustomEndpoint}

@@

The different wellness programs mentioned above all have a portal with a dashboard where all the participating employees can view their progress and compare it with their colleagues. The employees can also be organized into teams and specific team goals can be set for the teams. Administrators of a wellness program can measure the progress of the employees and their amounts of healthy activities through an administrative dashboard where statistics and summaries about the programs can be generated and viewed.

@@

\begin{figure}[!hbt]
\centering
\begin{chronology}[1]{2012}{2016}{10cm}[10cm]
\event{\decimaldate{20}{12}{2013}}{Firbit}
\event{\decimaldate{12}{10}{2014}}{Jawbone}
\event{\decimaldate{5}{1}{2015}}{Garmin}
\end{chronology}
\caption{A timeline showing when three different smart wear manufacturers released their wellness programs\cite{fitbitWellnessStat}\cite{jawboneWellnessStat}\cite{garminWellnessStat}.}
\label{fig:timline}
\end{figure}

@@

Fruit Street is a software system built to help health care providers connect with their patients and is made for the American market. Fruit Street makes it possible for the clinicians to have video chats and screen sharing sessions with their patients. Appointments can be booked and payments collected. Patients can upload pictures of what food they are eating and share it with the clinicians for nutrition purposes.

@@

The user-interface to be used by the physiotherapist should be easy to use. It was also expressed that there was a need for at least two kinds of views: (1) An overview, where all patients can be listed and where it is easy to see the status of each patient. (2) A specific view where details about each patient can be viewed.

@@

Two things had to be taken into account when the requirements for RSAT were made. (1) Sony Lifelog was already the chosen data source. This meant that only data that could be tracked by Lifelog devices and that was accessible via the Lifelog API could be fetched into RSAT.  (2) RSAT was to be integrated with a system that already existed (the itACiH planning system). This meant that the architecture of RSAT had to be designed such that an integration was possible.

@@

\section{Requirements}
\label{sec:req}
%TODO: Skriva ingress här? Kanske iaf...

@@

It was also decided that it should be possible to browse the history of all the collected data in RSAT. This function was considered important, since it would make it possible to see if patients tracked in RSAT makes any progress, i.e. if they increase their amount of physical activity or not.

@@

\begin{itemize}
    \item RSAT shall be built as a system the can be reached from the Internet.
    \item The user-interface shall be easy to use.
    \item The data in RSAT shall be easy to access for the staff using it.
    \item It shall be possible to create specific goals for each patient in the system.
    \item It shall be possible to create goals for each datatype that exists in RSAT.
    \item It shall be possible for the clinician to have an overview of all patients, showing if they have completed their goals or not.

@@

\subsection{Other requirements}
\begin{itemize}
    \item Each patient shall be able to view the goals that the clinician has created for him/her.
    \item It shall be possible to see if a patient has been inactive for too long.
    \item It shall be possible to integrate other wearable devices from other manufacturers in future versions of RSAT.
    \item A device/set of devices (smartphone and wristband) used by a patient shall be reusable. A new patient shall be able to be equipped with the same device/set of devices that have been used by another patient before.
\end{itemize}

\section{Initial idea}
The initial idea was to develop a specific Android application that was installed on each patient’s smartphone. This would have been the ultimate choice but this idea was thrown and it was chosen that the patients would use the Sony Lifelog Application instead. The two major reasons for this decision were: (1) Sony has invested a lot of resources to create the Lifelog application that runs on most newer Android devices. This application has almost all the functionality that the physiotherapist initially asked for.
(2) That the schedule for building the RSAT was quite tight. If an Android application should be developed the time frame would have to be changed. It would also be hard to create an application that the patient’s would prefer to use instead of the Lifelog application.

The Sony Lifelog API is a read-only API. Hence, it is not possible to write any data to the API. It is therefore not possible for the physiotherapist to set goals that are automatically updated in the patients Lifelog accounts. Another problem is that the goals are not accessible at all through the API. This means that in order to set goals, the physiotherapist must have a meeting with the patient where they together set the different goals for the patient. At this meeting, the physiotherapist inserts the goals in RSAT and the patient inserts the goals in the Lifelog application.

\section{Final idea}
\label{sec:proposed-solution}
% mer lim
It was decided that the patients tracked in RSAT will use the Lifelog application. It was also decided that a web application will be developed that the clinicians can use to track the activities of the patients.

Figure \ref{fig:pro} shows an overview of the solution followed by a description of the steps for transferring the data from the patient to the physiotherapist.

@@

\begin{enumerate}
    \item Each patient added in RSAT has a smart Sony SmartBand wristband around his or her wrist.
    \item The wristband collects data about each patient's movements and the data is categorized into different activities.
    \item A Sony smartphone collects the data from the wristband. 
    \item The data is stored in the Sony Lifelog application on the smartphone and synchronized to Sony's cloud.
    \item RSAT fetches data from each patient's RSAT-account in the Lifelog cloud, processes it and saves it to a database.
\item The RSAT web application retrieves the data from all patients in the database and visualizes it to the clinicians.
\item The clinicians can interact with RSAT and see the progress of each patient via the web application. 
\end{enumerate}

@@

Each patient in the program will be equipped with a Sony SmartBand and an Android smartphone that are borrowed from the hospital. The SmartBand must synchronize its data to a smartphone and that is the reason why the patients are also equipped with a smartphone and not only a SmartBand.

To make it easy for the clinicians and the patients to get started, each smartphone and SmartBand will be  preconfigured by an administrator. The following steps are used to configure the devices and accounts for the patients:

\begin{enumerate}
    \item{A new Google account is created to be used in the Lifelog application.}
    \item{All neccessary applications are downloaded to the smartphone.}
    \item{The SmartBand is connected with the smartphone.}
    \item{The smartphone is added to RSAT and given a unique name.}
\end{enumerate}

When a new patient is added to RSAT, the clinician gives him/her the preconfigured set of devices. The clinican also links the patients profile to the smartphone's name in RSAT.

\chapter{Development}
This chapter describes the development cycles of the RSAT. It also gives a brief introduction to the different 3rd-party libraries and PalCom services that have been used to build RSAT.

\section{3rd-party libraries and services}

@@

% Lägga till användare, bökigt i vårt system...

@@

% PRATA OM ANGEL...
